---
title: "R for Data Science"
subtitle: "Set up a project, then play the *whole game*: load → tidy → transform → visualise"
author: "Postgraduate R Tutorial"
date: today
format:
  revealjs:
    embed-resources: true
    theme: [default, custom.scss]
    slide-number: c/t
    show-slide-number: all
    incremental: false
    code-line-numbers: true
    code-overflow: wrap
    highlight-style: github
    footer: "Adapted from *R for Data Science (2e)*, Wickham, Çetinkaya-Rundel & Grolemund — r4ds.hadley.nz"
    width: 1280
    height: 760
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
  fig-width: 7
  fig-height: 4.2
  dpi: 150
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(palmerpenguins)
theme_set(theme_minimal(base_size = 14))
options(pillar.print_max = 6, pillar.print_min = 6)
```

# Welcome {.center}

::: {style="font-size:0.9em"}
Today we do two things:

1. **Set up a project properly** — folders, paths, and the idea of *tidy data* — so your
   work is organised and reproducible from day one.
2. Then play the **whole game** of data science, following data in its natural direction:
   **load it in → tidy it → transform it → visualise it.**
:::

::: {.notes}
This mirrors Part 1 ("Whole game") of R4DS, but reordered around the real workflow: get
organised first, then move data through the pipeline from import to plot.
:::

## The data science cycle

> **Import → Tidy → Transform ⇄ Visualise → (Model → Communicate)**
> ...all sitting inside **Program**.

. . .

We'll follow it left to right today:

- **Set up** — a project structure that keeps everything findable and reproducible.
- **Import** — get a data file into R.
- **Tidy** — reshape it into a consistent form.
- **Transform** — filter, derive variables, summarise by group.
- **Visualise** — turn the table into a picture. *(We finish here.)*

# Part 1 · Set up your project {.center}

> Get organised *before* you analyse. Future-you will thank present-you.

## Always work in an RStudio Project

An **RStudio Project** (`.Rproj`) bundles everything for one piece of work.

- Open the project and your **working directory is the project folder** — automatically.
- No more `setwd("C:/Users/neil/Desktop/...")` that breaks on every other machine.
- Each project gets its own history, and you can have several open at once.

. . .

> Create one with **File → New Project → New Directory**.
> One project = one paper / thesis chapter / analysis.

::: {.notes}
The Project is the unit of reproducibility. Everything below assumes you're inside one.
:::

## A sensible folder structure

```text
my-analysis/
├── my-analysis.Rproj      # the project file — open THIS
├── README.md              # what is this, who, how to run it
├── data-raw/              # original data — READ-ONLY, never edit by hand
├── data/                  # cleaned / derived data you generate from data-raw
├── Code/                  # your analysis scripts (.R) or Quarto docs (.qmd)
├── output/
│   ├── figures/           # plots you export
│   └── tables/            # results tables
└── renv/                  # optional: a reproducible package library
```

. . .

Two principles: **raw data is sacred and read-only**; **everything in `output/` is
regenerable** by re-running your scripts.

## Paths that don't break

```{r}
#| eval: false
# Bad — absolute, machine-specific:
read_csv("C:/Users/neil/Desktop/project/data-raw/palmerpenguins.csv")

# Good — relative to the project root:
read_csv("data-raw/palmerpenguins.csv")

# Robust — builds the path correctly on any OS / machine:
library(here)
read_csv(here("data-raw", "palmerpenguins.csv"))
```

Because you're in a Project, relative paths just work. `here::here()` makes them
bullet-proof even from sub-folders.

## Reproducible habits

- Write your analysis in a **script** (`.R`) or **Quarto** (`.qmd`) — not the console — so it's saved and re-runnable.
- Run a line / selection with **Cmd/Ctrl + Enter**; run the whole file with **Cmd/Ctrl + Shift + Enter**.
- **Restart R often** (Session → Restart R) and re-run from the top.

. . .

> Turn off "save workspace to .RData": **Tools → Global Options → uncheck "Restore .RData"**.
> A script that only works because of leftover objects isn't reproducible.

# Part 2 · R foundations {.center}

> The minimum syntax you need before touching data.

## Assignment & naming

```{r}
#| eval: false
# Assign with <-   (shortcut: Alt/Option + -)
mean_mass <- 4200

# Names: lowercase, words separated by _ (snake_case)
flipper_summary <- 1   # good
FlipperSummary  <- 1   # avoid
flipper.summary <- 1   # avoid
```

- Use `<-` to assign; reserve `=` for arguments inside functions.
- Nothing prints when you assign — type the object's name to see it.

## Calling functions

```{r}
seq(from = 1, to = 10, by = 2)
```

- Functions take **arguments** as `name = value`.
- First arguments are usually passed by position; name the rest for clarity.
- `?seq` opens the help page; Tab-completion shows argument names.

::: {.notes}
Demo tab-completion and matching-parenthesis highlighting. Unclosed quotes/parens give a
`+` continuation prompt — press Esc to escape.
:::

## Code style

```{r}
#| eval: false
# Spaces around operators and after commas:
mean(x, na.rm = TRUE)      # good
mean(x,na.rm=TRUE)         # cramped

# One statement per line; descriptive snake_case names:
mean_body_mass <- mean(penguins$body_mass_g, na.rm = TRUE)
```

- Spaces around `<-`, `==`, `+`, `*`; a space after every comma.
- The **{styler}** package reformats code automatically; **{lintr}** flags problems.

> Code is read far more often than written — including by you, in six months.

# Part 3 · Tidy data {.center}

> The shape we're aiming for. Get here and everything downstream is easy.

## The three rules of tidy data

::: {.cols}
::: {.col}
1. Each **variable** is a column.
2. Each **observation** is a row.
3. Each **value** is a cell.
:::
::: {.col}
For us: **one row per specimen**, one column per measured trait, with `species`, `site`,
and `sex` as their own columns.

Tidy data is what `dplyr`, `ggplot2` and most models expect. Field data rarely arrives
this way — Part 4 shows how to get it there.
:::
:::

# The whole game {.center}

> **Import → Tidy → Transform → Visualise.** Now we run real data through it.

## Import: `read_csv()`

```{r}
#| eval: false
library(tidyverse)

penguins <- read_csv("data-raw/palmerpenguins.csv")
```

. . .

`readr::read_csv()` gives you:

- a **tibble** (a tidy-friendly data frame),
- automatic **column-type guessing**, printed so you can check it,
- handy arguments: `na = c("", "NA", "-999")`, `skip =`, `col_types =` to override guesses.

> Always eyeball the type report and first rows — silent mis-typing is the #1 import bug.

## Other importers

| Format | Function | Package |
|---|---|---|
| `.csv` / delimited | `read_csv()`, `read_delim()` | readr |
| Excel `.xlsx` | `read_excel()` | readxl |
| `.rds` (one R object) | `read_rds()` | readr |
| SPSS / Stata / SAS | `read_sav()`, `read_dta()` | haven |

## Meet the data

We'll use **morphometric data** — measurements across species and sites. Once loaded:

```{r}
penguins
```

## A first look

```{r}
glimpse(penguins)
```

`glimpse()` is your first move on any new dataset: every column, its type, and a peek at
the values.

## Tidy in practice: a messy table

Field data often arrives **wide** — species as column headers:

```{r}
counts_wide <- tribble(
  ~site,   ~Adelie, ~Gentoo, ~Chinstrap,
  "North",      12,      30,          5,
  "South",      18,      22,          9
)
counts_wide
```

`species` is a *variable* trapped in the *headers* — not tidy.

## `pivot_longer()` to the rescue

```{r}
#| code-line-numbers: "2-5"
counts_long <- counts_wide |>
  pivot_longer(
    cols      = -site,
    names_to  = "species",
    values_to = "count"
  )
counts_long
```

Now `species` and `count` are proper columns — ready to group, plot, model.
(`pivot_wider()` does the reverse.)

## Transform: the pipe `|>`

The pipe passes the result on the left into the first argument on the right.

```{r}
#| eval: false
# Nested (hard to read):
summarise(group_by(filter(penguins, !is.na(sex)), species), n = n())

# Piped (read top to bottom):
penguins |>
  filter(!is.na(sex)) |>
  group_by(species) |>
  summarise(n = n())
```

> Read `|>` as **"and then"** — it turns nested calls into a readable recipe.

## `filter()` — keep rows

```{r}
penguins |>
  filter(species == "Adelie", body_mass_g > 4000)
```

Operators: `==  !=  <  >  <=  >=  &  |  %in%`, plus `is.na()`.

## `arrange()` & `select()`

::: {.cols}
::: {.col}
**Reorder rows**
```{r}
penguins |>
  arrange(desc(body_mass_g)) |>
  select(species, body_mass_g)
```
:::
::: {.col}
**Pick columns**
```{r}
penguins |>
  select(species, island, sex)
```
:::
:::

## `mutate()` — create variables

```{r}
#| code-line-numbers: "2-3"
penguins |>
  mutate(
    mass_kg    = body_mass_g / 1000,
    bill_ratio = bill_length_mm / bill_depth_mm
  ) |>
  select(species, mass_kg, bill_ratio)
```

Where derived morphometrics live — ratios, indices, log-transforms for allometry.

## `group_by()` + `summarise()`

The workhorse for "compute something per species / site / treatment":

```{r}
penguins |>
  group_by(species, sex) |>
  summarise(
    n         = n(),
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass   = sd(body_mass_g, na.rm = TRUE),
    .groups   = "drop"
  )
```

::: {.notes}
Flag na.rm = TRUE — one NA poisons the mean otherwise. .groups = "drop" avoids the
"still grouped" surprise downstream.
:::

# Visualise {.center}

> The payoff. We finish the game by turning data into pictures.

## The grammar of graphics

Every ggplot is built from the same pieces:

```r
ggplot(data = <DATA>) +
  geom_<TYPE>(mapping = aes(<MAPPINGS>))
```

. . .

- **data** — the data frame.
- **aes()** — *aesthetic mappings*: which variable maps to x, y, colour, shape…
- **geom_** — the geometric object representing the data (points, lines, bars…).
- Layers are added with **`+`**.

## A first plot

```{r}
#| label: first-plot
#| output-location: column-fragment
#| fig-width: 5.4
#| fig-height: 4.6
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm,
                y = body_mass_g)
) +
  geom_point()
```

::: {.notes}
Build it live: ggplot(penguins) -> blank canvas, add aes, then geom_point. A blank grey
panel is "correct" until you add a geom.
:::

## Map a variable to colour

```{r}
#| label: colour-plot
#| output-location: column-fragment
#| fig-width: 5.4
#| fig-height: 4.6
#| code-line-numbers: "4"
ggplot(
  penguins,
  aes(x = flipper_length_mm,
      y = body_mass_g,
      colour = species)
) +
  geom_point()
```

Anything that should respond to the data goes **inside `aes()`**.

## Set vs. map

Want one fixed colour? **Set** it *outside* `aes()`:

```{r}
#| label: set-aes
#| output-location: column-fragment
#| fig-width: 5.4
#| fig-height: 4.6
#| code-line-numbers: "2"
ggplot(penguins, aes(flipper_length_mm, body_mass_g)) +
  geom_point(colour = "steelblue", alpha = 0.7)
```

> **Map** = a variable controls it (inside `aes()`). **Set** = fixed constant (outside `aes()`).

## Add a model layer

```{r}
#| label: smooth-plot
#| output-location: column-fragment
#| fig-width: 5.4
#| fig-height: 4.6
#| code-line-numbers: "7"
ggplot(
  penguins,
  aes(x = flipper_length_mm,
      y = body_mass_g,
      colour = species)
) +
  geom_point() +
  geom_smooth(method = "lm")
```

One trend line **per group**, because `colour` is mapped globally.

## Facets — small multiples

```{r}
#| label: facet
#| fig-width: 9
#| fig-height: 3.6
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point() +
  facet_wrap(~ island)
```

`facet_wrap(~ var)` splits one plot into a panel per level — ideal for sites / treatments.

## Putting it together

The whole game in one piped pipeline, ending in a plot:

```{r}
#| label: capstone
#| fig-width: 8
#| fig-height: 3.3
penguins |>                                   # import (loaded earlier)
  filter(!is.na(body_mass_g), !is.na(sex)) |> # transform: drop missing
  mutate(mass_kg = body_mass_g / 1000) |>     # transform: derive variable
  group_by(species, sex) |>                   # transform: group
  summarise(mean_kg = mean(mass_kg), .groups = "drop") |>
  ggplot(aes(species, mean_kg, fill = sex)) + # visualise
  geom_col(position = "dodge") +
  labs(x = NULL, y = "Mean mass (kg)", fill = "Sex")
```

# Wrap up {.center}

## What we covered

::: {.cols}
::: {.col}
**Set-up & foundations**

- RStudio Projects + folder structure
- Relative paths / `here()`
- Assignment, naming, functions, style
- Tidy data: the 3 rules
:::
::: {.col}
**The whole game**

- `read_csv()` and friends
- `pivot_longer()` → tidy
- dplyr: filter / arrange / select / mutate / summarise + `|>`
- ggplot2: aes, geoms, set vs map, facets
:::
:::

## Your turn — exercises

1. Make a project folder for a small analysis with `data-raw/`, `data/`, `R/`, and `output/figures/`.
2. Read **one of your own CSVs** with `read_csv()`, check the column types, and `glimpse()` it.
3. Compute the number of specimens and mean body mass **per island** (`group_by` + `summarise`).
4. Plot `bill_length_mm` vs `bill_depth_mm`, colour by `species`, add `geom_smooth(method = "lm")`. Then facet by species — what changes? *(Simpson's paradox!)*

::: {.muted .small}
Bonus: take a wide spreadsheet from your own work and tidy it with `pivot_longer()`.
:::

## Getting unstuck

- **`?function`** — the help page (the Examples section is gold).
- **`vignette("dplyr")`** — narrative, worked guides for a package.
- Google the **exact error message**; Posit Community & Stack Overflow `[r]` `[tidyverse]`.
- Make a **reprex** with the **{reprex}** package before asking — you'll often solve it yourself.

## Where to go next

- 📖 **R for Data Science (2e)** — free: [r4ds.hadley.nz](https://r4ds.hadley.nz)
- 🧾 **Posit cheatsheets** — ggplot2, dplyr, tidyr: [posit.co/resources/cheatsheets](https://posit.co/resources/cheatsheets)
- 🐞 For ecology, pair these with `{vegan}`, `{lme4}`/`{glmmTMB}`, `{DHARMa}`

. . .

> The whole game is just this loop, run over and over:
> **import → tidy → transform ⇄ visualise.** Everything else is depth.

## Thank you {.center}

Questions?

::: {.muted .small}
Slides built with Quarto + reveal.js. Source: `whole-game.qmd`.
Content adapted from *R for Data Science (2e)* (CC BY-NC-ND).
:::
